A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-11 doc/popl2013/evaluation.tex

We evaluate the Legion type system on three criteria:
\begin{itemize}
\item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
\item Overhead: Does it reduce checking costs (Section~\ref{subsec:overhead})?
\item Scalability: Can it leverage hierarchical scheduling (Section~\ref{subsec:scalability})?
\end{itemize}
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 processors, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

@@

\subsubsection{Fluid Simulation}
\label{subsec:fluid}
Fluid is based on the {\tt fluidanimate} benchmark from the PARSEC 
suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
using particles that move within a regular grid of cells.  To perform operations in 
parallel the cells are partitioned.  Unlike
Circuit, Fluid first creates regions and partitions them before
allocating cells in regions.  Another difference is that Fluid
maintains separate regions for ghost cells rather using multiple partitions of
the regions containing shared data.  Region relationships are used for remembering
which regions are required for each grid.

\subsubsection{Adaptive Mesh Refinement}
\label{subsec:amr}
The third application is an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with subrefinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
The AMR application uses separate regions for every level of
refinement.  The regions at each level are partitioned several ways to provide multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces at 
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions can be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

These applications show Legion is capable of 
expressing applications with both regular (Fluid,AMR) and irregular (Circuit) pointer data structures.
Being able to dynamically create and partition regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization(AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit,AMR).  All the types of privileges and coherence we have
described are needed in different applications.  Region relationships
are used in all applications. %, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

\subsection{Checking Overhead}
\label{subsec:overhead}
Our first Legion implementation consisted of a C++ library of Legion primitives
\cite{Legion12} with no checking of region memory accesses.  
When testing applications we frequently encountered memory corruption due to
illegal region accesses caused by a confluence of application bugs and lack of checking.  Many times these corruptions 
occurred on remote nodes or on GPUs, requiring heroic debugging efforts.
%  Finding these bugs required either multiple {\tt gdb} sessions 
%for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
%GPUs.  
To mitigate this problem and guarantee safety we added 
dynamic checks on  all region accesses for both CPUs and GPUs.  However, these checks
added considerable runtime overhead.  
To avoid the overhead of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
We rewrote all of our applications in this language and used our type checker to verify them.  Once the applications type checked we were
able to safely elide all the dynamic region access checks.

@@

In addition to total processor overhead, we also measured performance gain from eliding checks in 
terms of wall-clock time.  Since most region accesses occur in leaf tasks the checks parallelize 
well.  Wall-clock performance gains from eliding memory checks ranged from 1-10\%, 1-15\%, 
and 2-71\% for Circuit, Fluid, and AMR respectively.  Performance gains for AMR were larger than
the other applications because AMR was already memory bound and the additional checks intensified
memory pressure.  For the GPU kernels in the Circuit application checking required up to 8 additional 
registers per thread.  While the GPU kernels in Circuit where not bound by 
available on-chip memory, kernels that are would be susceptible to extreme performance 
degradation due to the extra registers required for checking.  We also measured the overhead
of the dynamic checks associated with checking task call privileges but found them to be negligible.
% relative to the overhead of region access checks.

\subsection{Scalability}
\label{subsec:scalability}
The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
theorem makes it safe for the runtime to only perform dependence checks between two
tasks that share the same immediate parent task instead of having to consider all
possible interactions with other tasks anywhere in the system.  Since all tasks that
share the same immediate parent are all created on the node where the parent is executing,
all dependence checks and scheduling decisions can be performed locally with no communication.  
Without this property the runtime would be forced to inform all other processors in the system 
of every task that it launches using expensive broadcasts.

>>>>

We evaluate the Legion type system on three criteria:
\begin{itemize}
\item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
\item Overhead: Does it reduce checking costs (Section~\ref{subsec:overhead})?
\item Scalability: Can it leverage hierarchical scheduling (Section~\ref{subsec:scalability})?
\end{itemize}
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

@@

\subsubsection{Fluid Simulation}
\label{subsec:fluid}
Fluid is based on the {\tt fluidanimate} benchmark from the PARSEC 
suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
using particles that move within a regular grid of cells.  To perform operations in 
parallel the array of cells is partitioned.  Unlike
Circuit, Fluid first creates regions and partitions them before
allocating cells in regions.  Another difference is that Fluid
maintains separate regions for ghost cells rather using multiple partitions of
the regions containing shared data.  Region relationships are used for remembering
which regions are required for each grid.

\subsubsection{Adaptive Mesh Refinement}
\label{subsec:amr}
The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with subrefinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
Every levelof refinement uses a separate region, which is partitioned several ways to support multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces at 
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

These applications show Legion is capable of 
expressing applications with both regular (Fluid, AMR) and irregular (Circuit) pointer data structures.
Being able to dynamically create and partition regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence we have
described are needed in different applications.  Region relationships
are used in all applications. %, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

\subsection{Checking Overhead}
\label{subsec:overhead}
Our first Legion implementation consisted of a C++ library of Legion primitives
\cite{Legion12} with no checking of region memory accesses.  
When testing applications we frequently encountered memory corruption due to
illegal region accesses caused by application bugs.
In many cases, this corruption occurred between nodes in the cluster or on 
GPUs, environments for which debugging tools are primitive at best.
%  Finding these bugs required either multiple {\tt gdb} sessions 
%for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
%GPUs.  
To location the application bugs causing these illegal accesses, we initially added 
dynamic checks on  all region accesses for both CPUs and GPUs.  
Unsurprisingly, these checks added considerable runtime overhead.  
To preserve the benefit of checking every access, but eliminate the cost of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
We then rewrote all of our applications in this language and used our type checker to verify them.  Once the applications type checked we were
able to safely elide all the dynamic region access checks.

@@

In addition to total processor overhead, we also measured performance gain from eliding checks in 
terms of wall-clock time.  Since most region accesses occur in leaf tasks the checks parallelize 
well.  Wall-clock performance gains from eliding memory checks ranged from 1-10\%, 1-15\%, 
and 2-71\% for Circuit, Fluid, and AMR respectively.  Performance gains for AMR were larger than
the other applications because AMR was already memory bound and the additional checks intensified
memory pressure.  For the GPU kernels in the Circuit application checking required up to 8 additional 
registers per thread.  While the GPU kernels in Circuit were not bound by 
available on-chip memory, kernels that are would be susceptible to extreme performance 
degradation due to the extra registers required for checking.  We also measured the overhead
of the dynamic checks associated with checking task call privileges but found them to be negligible.
% relative to the overhead of region access checks.

\subsection{Scalability}
\label{subsec:scalability}
The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
theorem proves it is safe for the runtime to only perform dependence checks between two
tasks that share the same immediate parent task instead of having to consider all
possible interactions with other tasks anywhere in the system.  Since all tasks that
share the same immediate parent are all created on the node where the parent is executing,
all dependence checks and scheduling decisions can be performed locally with no inter-processor communication.  
Without this property the runtime would be forced to inform all other processors in the system 
of every task that it launches using expensive broadcasts.
